Stage-Splitting a Modal Language

### 2014-08-29 paper/semantics.tex
added more about a partial evaluation system

%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
The language \lang\ is essentially two copies of a typed lambda calculus with
products, sums, and isorecursive types.%
\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
expressing in which copy of the language the term resides. Intuitively, the
stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.

@@

\subsection {A Partial Evaluation System}

We now consider how to use this formal system to encode inputs to partial evaluation and stage-splitting systems.
Recall that in partial evaluation systems, we have a function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, we can partially evaluate the function to produce
a residual that depends on only the dynamic input.

By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
we can encode this sort of system with a function type of the form $A\to\fut(B\to C)$.
\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
the following higher-order function which makes the adjustment:
\begin{lstlisting} 
let adjust (f : $int * int -> $int) =
  fn (p : int) => 
    next{
      fn (b : int) => 
        prev{f (next {b}, p)}
    }
\end{lstlisting}}
%
Here $A$ represents the static input, $B$ represents the dynamic input, and $C$ represents the output.

>>>>

\lang\ is largely based off of a language defined in \cite{davies96} 
which was used as a model for partial evaluation.  
Our version, despite a slightly different semantics, can be used for the same purpose.
The details of the partial evaluation view are shown in \cref{sec:partialeval}.

The language \lang\ is essentially two copies of a typed lambda calculus with
products, sums, and isorecursive types.%
\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
expressing in which copy of the language the term resides. Intuitively, the
stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.

@@

\subsection {A Partial Evaluation System}
\label{sec:partialeval}

We now consider how to use this formal system to encode inputs to partial evaluation and stage-splitting systems.
Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, we can partially evaluate the function to produce
a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is residual, also called ``f specialized to x."
The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must evaluate it many times.

By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
our dynamics can serve as a partial evaluator.   
Specifically, we encode $f$ as a \lang\ expression with a function type of the form $A\to\fut(B\to C)$
\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
the following higher-order function which makes the adjustment:
\begin{lstlisting} 
let adjust (f : $int * int -> $int) =
  fn (p : int) => 
    next{
      fn (b : int) => 
        prev{f (next {b}, p)}
    }
\end{lstlisting}}.
%
Here $A$ represents the static input, $B$ represents the dynamic input, and $C$ represents the output.

### 2014-08-29 paper/splitting.tex
added more about a partial evaluation system

The rules for value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original value.  
At base types, value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
in accordance with the precept that base types have only trivial content at stage \bbtwo.
Splitting essentially distributes into values at product, sum, and recursive types.
Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 
general term splitting, which is covered momentarily.

>>>>

The rules for partial value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original partial value.  
At base types, partial value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
in accordance with the precept that base types have only trivial content at stage \bbtwo.
Splitting acts distributively at product, sum, and recursive types.
Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 
general term splitting, which is covered momentarily.
